The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-03-29 Paper.Rmd
further work on Discussion

<!--what we did and why is it important to study?-->
We showed that non-N copula structure and asymmetric tail dependence are 
exceedingly common in datasets that range across multiple sub-disciplines
and levels of organization in ecology, in spite of the fact that no published 
ecological research we are aware of has 
taken this structure into account. Via simple mathematical models, we hypothesized two 
mechanisms that may cause non-N copula structures with asymmetric tail dependence
in ecological data. One
mechanism was via a Moran-like effect by which the dependence structure between ecological
variables (measured through time at different locations) is shaped by the dependence structure of 
environmental variables (measured at the same times and locations). 
A second mechanism gives rise to non-N copula structure and asymmetric tail dependence via
the asymmetric or nonlinear influence of environmental variables on ecological variables, 
and relates to Liebig's law of the minimum. We discuss below how commonly these and other
potential mechanisms may opperate. 
<!--***DAN: make sure to do so-->
We also demonstrated important consequences of 
non-N and tail-dependent copula structure in ecology. First, the skewness, through time, 
of the spatial average of time series from different locations (e.g., local pest abundance
time series) is affected by the tail dependence of the relationships
between the local quantities. Thus right-tail-dependent local quantities 
tend to produce "spiky" spatially averaged time series, with large outbreaks. 
Second, metapopulation extinction risk can be accentuated or reduced by the tail dependence
properties of environmental variables that influence the populations. Finally, 
simulation results in an Appendix showed that spatial Taylor's law can be strongly influenced by
tail dependence between local populations. In our view, our results show that, for many
ecological phenomena, a complete understanding is impossible without 
characterizing dependence between variables using copulas.

@@

<!--***DAN: some mechanism paragraphs follow-->

@@

<!--A para that sets up Shya's extinction risk paper - asks about importance of this effect for real 
species and whether it needs to be incorporated into PVA, and in what cases. If the copula structure of environmental variables changes (is it changing?) what will be the impact on extinction risk? Also, there may be copula structure between vital rates, and it is not accounted for in typical stochastic matrix model based PVA. Does it need to be?-->
Although we have demonstrated that tail dependence between environmental variables
can influence extinction risk, substantial work remains to be done regarding the 
circumstances under which this result holds, and its magnitude in different scenarios. 
First, we used 
a non-density dependent model to examine extinction risk; does tail dependence
in environmental variables still influence extinction risk when density dependence
plays a role? Second, we considered metapopulation extinction risk, but the whole field 
of population viability analysis [@REF-Morris and Doak] stochastic matrix modelling

<!--Maybe a para from Holder on his thoughts in the importance of copulas for
macroevolutionary biology-->

<!--Conclusion of some kind-->

@@

We thank the many contributors to the large datasets we have used; D. Stevens and P. Verrier for data extraction; and Joel E. Cohen, Jonathan Walter, Thomas Anderson, and Lei Zhao for helpful suggestions. We thank James Bell of the Rothamsted Insect Survey (RIS). 
The Rothamsted Insect Survey, a National Capability, is funded by the Biotechnology and Biological Sciences Research Council under the Core Capability Grant BBS/E/C/000J0200. SG, LWS and DCR were partly funded by US National Science Foundation grants 0075516 and 1442595 and the James S McDonnell Foundation.

>>>>

<!--what we did and why is it important to study?-->
We showed that non-N copula structure and asymmetric tail dependence are 
exceedingly common in datasets that range across multiple sub-disciplines
and levels of organization in ecology, in spite of the fact that no published 
ecological research we are aware of has 
taken this structure into account. Via simple mathematical models, we hypothesized two 
mechanisms that may cause non-N copula structures with asymmetric tail dependence
in ecological data. One
mechanism was via a Moran-like effect by which the dependence structure between ecological
variables (measured through time at different locations) is shaped by the dependence structure of 
environmental variables (measured at the same times and locations). 
A second mechanism gives rise to non-N copula structure and asymmetric tail dependence via
the asymmetric or nonlinear influence of environmental variables on ecological variables, 
and relates to Liebig's law of the minimum. We discuss below how commonly these and other
potential mechanisms may opperate. 
We also demonstrated important consequences of 
non-N and tail-dependent copula structure in ecology. First, the skewness, through time, 
of the spatial average of time series from different locations (e.g., local pest abundance
time series) is affected by the tail dependence of the relationships
between the local quantities. Thus right-tail-dependent local quantities 
tend to produce "spiky" spatially averaged time series, with large outbreaks. 
Second, metapopulation extinction risk can be accentuated or reduced by the tail dependence
properties of environmental variables that influence the populations. Finally, 
simulation results in an Appendix showed that spatial Taylor's law can be strongly influenced by
tail dependence between local populations. In our view, our results show that, for many
ecological phenomena, a complete understanding is impossible without 
characterizing dependence between variables using copulas.

@@

<!--some mechanism paragraphs follow-->

@@

<!--A para that sets up Shya's extinction risk paper-->
Although we have demonstrated that tail dependence between environmental variables
can influence extinction risk, substantial work remains to be done regarding the 
circumstances under which this result holds, and its magnitude in different scenarios. 
First, we used 
a non-density dependent model to examine extinction risk. Does tail dependence
in environmental variables still influence extinction risk when density dependence
plays a role? Second, we considered metapopulation extinction risk, but the large field 
of population viability analysis (PVA) via stochastic 
matrix modelling [@Caswell2000; @MorrisDoak2002]
uses a framework in which a single population's vital rates (e.g., life-stage-specific 
fecundity and survival rates) are considered to
vary stochastically through time due to environmental variation. Do relationships
between different vital rates exhibit tail dependence, and does tail dependence 
influence extinction risk of a single population?
PVA is one of the routes by which a species can be added to the influential
International Union for the Conservation of Nature (IUCN) red list of threatened
species. It seems worthwhile for future research to investigate the extent to which 
copula structure in dependences between vital rates may occur and may influence
estimates of risk. Finally, is the copula structure of environmental variables
changing through time, and, if so, how do such changes influence extinction risks?
Climate change is known to amplify the factors that lead to extreme weather events
[@Hansen2012]. Extreme events, by definition, produce values in the right tails of distributions
of environmental variables. Extreme weather events also tend to be spatially extensive, so
increased frequency of extreme weather events may correspond to changes in the tail dependence
environmental variables measured in different places, and we showed such changes
can affect extinction risk. It seems
important for future work to explore whether such effects are of sufficient 
commonness and magnitude to interest conservationists.

<!--Maybe a para from Holder on his thoughts in the importance of copulas for
macro-scale evolutionary biology?-->

<!--Brief conclusion of some kind-->

@@

We thank the many contributors to the large datasets we used; D. Stevens and P. Verrier for data extraction; and Joel E. Cohen, Jonathan Walter, Thomas Anderson, and Lei Zhao for helpful suggestions. We thank James Bell of the Rothamsted Insect Survey (RIS). 
The Rothamsted Insect Survey, a National Capability, is funded by the Biotechnology and Biological Sciences Research Council under the Core Capability Grant BBS/E/C/000J0200. SG, LWS and DCR were partly funded by US National Science Foundation grants 0075516 and 1442595 and the James S McDonnell Foundation.
